---
title: "Teaching Materials"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    self_contained: true
---


<!--
To compile run:
library(rmarkdown); render("teaching.Rmd")
To extract the R-code:
library(knitr); knitr::purl("teaching.Rmd")

HTS_introduction.pdf
DESeq2.pdf 
-->




```{r config, results='hide', include=FALSE}
path <- "http://www-huber.embl.de/users/klaus/EMBO.R"

```

## Overview 
    
This webpage contains teaching materials I have been using for courses. Each
topic is suitable for 3~6 hrs of an intensive course. The labs consist of
.Rnw or .Rmd files which require the installation of several packages. The necessary
packages for most of them can be  conveniently installed running the following script within R:

```{r , eval=FALSE}
## script location
script <- "http://www-huber.embl.de/users/klaus/EMBO.R"
## installation 
source(script)
```

However, the necessary packages are also indicated at the beginning of each lab.
The acknowledgments contain some references to the material used, which 
often represent links to additional documents suitable 
for further in-depth study of the topic.

Most of the labs have been built using
[knitr](http://yihui.name/knitr/) 
and [BiocStyle](http://bioconductor.org/packages/release/bioc/html/BiocStyle.html)
in connection with Sweave-like .Rnw documents.

## Introduction to R

This material gives a concise introduction to R. 

### Acknowledgements
Large parts of this material are based on the 
[contributed documentation](http://cran.r-project.org/other-docs.html)
on CRAN. Notably,
["Applied Statistics for Bioinformatics Using R"](http://cran.r-project.org/doc/contrib/Krijnen-IntroBioInfStatistics.pdf)
,
["IcebreakeR"](http://cran.r-project.org/doc/contrib/Robinson-icebreaker.pdf)
and
["A (very) short Introduction to R"](http://cran.r-project.org/doc/contrib/Torfs+Brauer-Short-R-Intro.pdf)
as well as the 
["Best first R tutorial"](https://www.nceas.ucsb.edu/files/scicomp/Dloads/RProgramming/BestFirstRTutorial.pdf)
and introductory material from [Laurent Gatto](https://github.com/lgatto/TeachingMaterial).


An in-depth resource for the details of R-programming is
[Advanced R](http://adv-r.had.co.nz/).

### Files

--------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Slides      [Slides](http://www.tinyurl.com/BK-XT85/Teaching/intro-R.pdf) 

Lab         [Lab pdf](http://www.tinyurl.com/BK-XT85/Teaching/R-lab.pdf)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/R-lab.R) 

Rnw         [.Rnw file](http://www.tinyurl.com/BK-XT85/Teaching/R-lab.Rnw)    
--------------------------------------------------------------------------------                                                                               

## Data handling with R

This material gives an introduction to data handling and data reshaping
with R, including a lot of data handling
techniques using the [dplyr](http://cran.r-project.org/package=dplyr) package
and reshaping using the [tidyr](http://cran.r-project.org/package=tidyr) and
[reshape2](http://cran.r-project.org/package=reshape2) packages. We also introduce
chaining with the [magrittr](http://cran.r-project.org/package=magrittr) package.

### Files

--------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Slides      [Slides](http://www.tinyurl.com/BK-XT85/Teaching/DataHandling.pdf) 

Lab         [Lab pdf](http://www.tinyurl.com/BK-XT85/Teaching/Data-Handling-lab.pdf)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/Data-Handling-lab.R) 

Rnw         [.Rnw file](http://www.tinyurl.com/BK-XT85/Teaching/Data-Handling-lab.Rnw)    
--------------------------------------------------------------------------------   

### Acknowledgements
The material on [dplyr](http://cran.r-project.org/package=dplyr) is to some extend
based on tutorials by [Kevin Markham](http://rpubs.com/justmarkham/dplyr-tutorial)
and [Dirk Schuhmacher](http://www.dirk-schumacher.net/). The illustration of
the dplyr verbs is adapted from a presentation of H. Wickham.

Example data were provided by Michele Christovao, Elisabeth Zielonka and Ina Kalinina.



## Exploratory data analysis and ggplot2

Typical summary statistics for location and scale as well as common diagnostic 
plots are presented. The plots are given both in base R as well as using
[ggplot2](http://cran.r-project.org/package=dplyr) commands.


### Acknowledgements

Large parts of this material are based on the plots used in
["Applied Statistics for Bioinformatics Using R"](http://cran.r-project.org/doc/contrib/Krijnen-IntroBioInfStatistics.pdf).

Lars Velten contributed the Protein exercise using ggplot. (Note the data
used in the protein-example has  been simulated and does not correspond to any real experimental data).

The ggplot explanations have been inspired by the
[ggplot2 book](http://www.springer.com/statistics/computational+statistics/book/978-0-387-98140-6)
as well as the 
[ggplot2 intro](http://jofrhwld.github.io/avml2012/) 
by [Josef Fruehwald](http://jofrhwld.github.io/).

[Wolfgang Huber](http://www.embl.de/research/units/genome_biology/huber/members/index.php?s_personId=CP-60002619)
provided nice thoughts & slides  on color usage for graphics.

### Files

--------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Slides      [Slides](http://www.tinyurl.com/BK-XT85/Teaching/EDA.pdf) 

Lab         [Lab pdf](http://www.tinyurl.com/BK-XT85/Teaching/EDA-lab.pdf)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/EDA-lab.R) 

Rnw         [.Rnw file](http://www.tinyurl.com/BK-XT85/Teaching/EDA-lab.Rnw)    
--------------------------------------------------------------------------------                                                                               


## Statistical methods for Bioinformatics 

This is a very concise introduction to important statistical methods in bioinformatics:
dimensionality reduction, clustering and regression.

These techniques are illustrated in the context of the analysis of (single cell)
RNA--Seq data.

### Acknowledgements

The  material on statistical distributions is based on
["Applied Statistics for Bioinformatics Using R"](http://cran.r-project.org/doc/contrib/Krijnen-IntroBioInfStatistics.pdf). 

I adapt the usage of the bodyfat data as an example data set for multivariate models
such as regression and PCA from
Michael Lavine's [Introduction to Statistical Thought](http://people.math.umass.edu/~lavine/Book/book.html), which is an excellent introductory statistics textbook in itself.

The slides are partially based on material by 
[Wolfgang Huber](http://www.embl.de/research/units/genome_biology/huber/members/index.php?s_personId=CP-60002619) and
[John Marioni](http://www.ebi.ac.uk/about/people/john-marioni) (EMBL-EBI).

### Files

--------------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Slides      [Slides](http://www.tinyurl.com/BK-XT85/Teaching/Regression_PCA_MDS_RNASeq.pdf) 

Lab         [Lab pdf](http://www.tinyurl.com/BK-XT85/Teaching/methods-lab.pdf)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/methods-lab.R) 

Rnw         [.Rnw file](http://www.tinyurl.com/BK-XT85/Teaching/methods-lab.Rnw) 
-----------------------------------------------------------------------------------


## Hypothesis Testing 

This is an introduction to hypothesis testing, including multiple testing
as well as advanced topics such as regularized t-statistics, independent
filtering and empirical null estimation.


### Acknowledgements

The  material on the basic tests is based on
["Applied Statistics for Bioinformatics Using R"](http://cran.r-project.org/doc/contrib/Krijnen-IntroBioInfStatistics.pdf). 

Almost all of the slides are by [Wolfgang Huber](http://www.embl.de/research/units/genome_biology/huber/members/index.php?s_personId=CP-60002619).

The material on tests for categorical data is mainly based on the book 
[Introductory Statistics with R](http://www.springer.com/mathematics/probability/book/978-0-387-79053-4), which is also great for learning R and statistics at the same time. 


### Files

--------------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Slides      [Slides](http://www.tinyurl.com/BK-XT85/Teaching/testing_Intro.pdf) 

Lab         [Lab pdf](http://www.tinyurl.com/BK-XT85/Teaching/Testing-lab.pdf)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/Testing-lab.R) 

Rnw         [.Rnw file](http://www.tinyurl.com/BK-XT85/Teaching/Testing-lab.Rnw)    
-----------------------------------------------------------------------------------


## Hypothesis Testing -- concise

This is a concise introduction to hypothesis testing, including only the most 
widely used tests but also explaining the idea of permutation tests, since these
can be very usfeul in pratctice.

The material also covers multiple testing
as well as advanced topics such as regularized t-statistics, independent
filtering and empirical null estimation.


### Acknowledgements

Ther permutation test explainations were inspired by [Tim Hesterberg's](http://www.timhesterberg.net/)
[excellent review on resampling](http://www.timhesterberg.net/bootstrap#TOC-What-Teachers-Should-Know-about-the-Bootstrap:-Resampling-in-the-Undergraduate-Statistics-Curriculum) for undergraduates.


### Files

--------------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Slides      [Slides](http://www.tinyurl.com/BK-XT85/Teaching/testing_Intro.pdf) 

Lab         [Lab pdf](http://www.tinyurl.com/BK-XT85/Teaching/testing-lab-concise.pdf)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/testing-lab-concise.R) 

Rnw         [.Rnw file](http://www.tinyurl.com/BK-XT85/Teaching/testing-lab-concise.Rnw)    
-----------------------------------------------------------------------------------



## Differential expression analysis of RNA-Seq data with DESeq2

This material introduces a complete workflow for DE analysis of RNA--Seq data
starting from the raw FASTQ files. It performs a re-analysis of the 
RNA-Seq data analyzed in 
Uslu et. al. -- 
[Long-range enhancers regulating Myc expression are required for  normal facial morphogenesis](http://dx.doi.org/10.1038/ng.2971), 2014


### Acknowledgements

The material is largely based on the documentation of the [DESeq2](http://bioconductor.org/packages/release/bioc/html/DESeq2.html)
package on Bioconductor by [Mike Love](http://www.mike-love.net/), 
[Simon Anders](http://www.embl.de/research/units/genome_biology/huber/members/index.php?s_personId=CP-60002977)
and [Wolfgang Huber](http://www.embl.de/research/units/genome_biology/huber/members/index.php?s_personId=CP-60002619).

The first part of the lab, from FASTQ files to the count-table follows
Anders et. al. 
- [Count-based differential expression analysis of RNA sequencing data using R and Bioconductor](http://dx.doi.org/10.1038/nprot.2013.099), 2013 closely.

[Simon Anders](http://www.embl.de/research/units/genome_biology/huber/members/index.php?s_personId=CP-60002977) also provided the slides.

-----------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Slides      [Slides DESeq2 - Wolfgang Huber](http://www.tinyurl.com/BK-XT85/DESeq2-Wolfgang.pdf)

Slides      [Slides DESeq2](http://www.tinyurl.com/BK-XT85/Teaching/DESeq2.pdf)

Slides      [Slides HT Sequencing](http://www.tinyurl.com/BK-XT85/Teaching/HTS_introduction.pdf) 

Lab         [Lab pdf](http://www.tinyurl.com/BK-XT85/Teaching/DESeq2-Analysis.pdf)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/DESeq2-Analysis.R) 

Rnw         [.Rnw file](http://www.tinyurl.com/BK-XT85/Teaching/DESeq2-Analysis.Rnw)    
------------------------------------------------------------------------------------


## Differential expression analysis of RNA-Seq  -- Predoc Course 2014

This material introduces a  workflow for DE analysis of RNA-Seq data
starting from the gene count table. It is similar to the worklflow above and
performs a re--analysis of the 
RNA-Seq data analyzed in 
Uslu et. al. -- 
[Long-range enhancers regulating Myc expression are required for  normal facial morphogenesis](http://dx.doi.org/10.1038/ng.2971), 2014

It has been created for the DNA/RNA module of the [2014 EMBL predoc course](http://www.embl.de/predoccourse/2014/modules/DNA_RNA_Biology/index.html) and uses html instead of LaTeX.

### Acknowledgements

The material is largely based on the documentation of the [DESeq2](http://bioconductor.org/packages/release/bioc/html/DESeq2.html)
package and the [rnaseqGene](http://bioconductor.org/help/workflows/rnaseqGene/) workflow on Bioconductor by [Mike Love](http://www.mike-love.net/), 
[Simon Anders](http://www.embl.de/research/units/genome_biology/huber/members/index.php?s_personId=CP-60002977)
and [Wolfgang Huber](http://www.embl.de/research/units/genome_biology/huber/members/index.php?s_personId=CP-60002619). 


-----------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Lab         [Lab html](http://www.tinyurl.com/BK-XT85/Teaching/DESeq2Predoc2014.html)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/DESeq2Predoc2014.R) 

Rmd         [.Rmd file](http://www.tinyurl.com/BK-XT85/Teaching/DESeq2Predoc2014.Rmd)    
------------------------------------------------------------------------------------

## Machine Learning  -- Predoc Course 2014

This is an introduction to Machine Learning, it is still work in progress.


### Acknowledgements
This is heavily based on material form S. Arora ([Bioc Seattle, Oct 14](http://bioconductor.org/help/course-materials/2014/SeattleOct2014/)) and VJ Carey ([Brixen 2011](http://bioconductor.org/packages/release/bioc/vignettes/MLInterfaces/inst/doc/MLprac2_2.pdf)).

-----------------------------------------------------------------------------------  
Type                Link 
------      ------------------
Lab         [Lab html](http://www.tinyurl.com/BK-XT85/Teaching/ML-lab.html)                             

R script    [R--script](http://www.tinyurl.com/BK-XT85/Teaching/ML-lab.R) 

Rmd         [.Rmd file](http://www.tinyurl.com/BK-XT85/Teaching/ML-lab.Rmd)    
------------------------------------------------------------------------------------




<!--
ffff
-->
